Towards Semantic Patches

### 2010-06-18 papers/MIPS2010/abstract.tex
factorisation de l'intro, introduction du langage kernel
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@107 403e655e-980a-47bd-8c60-1ee5348bb6b7

It may be fruitful to investigate the possible relationship between the
very \emph{methodologies} or even \emph{daily workflow} employed in
both fields of mathematics and computer science: one trying to prove a
theorem, the other constructing a program to fulfill a task. How does
one or the other elaborates his object of study? What kind of \emph{a
  posteriori} modification is he prone to doing? What do these
modifications imply on the validity of the whole edifice or the other
way around, how do one rely on existing work to build up new results?
% Y: Je ne comprends pas tout à fait la dernière question: 
Finally, how to relate how both scientists collaborate in a team?

Recent efforts in the formalization of mathematical results have
naturally led to these questions and many of them remain largely
unanswered, but the tendency seems to be to adapt existing methods
coming from software development, as illustrated for example by the
recent introduction of modules, file-based scripts and separate
compilation in proof assistants like \textsf{Coq}~{\cite{CoqDocWeb}}
or \textsf{Matita}~{\cite{Asperti06userinteraction}}, the use of
dependency management tools (\textsf{Make}) or version control system
(\textsf{GIT}, \textsf{Subversion}) to build and manage versions of a
project. Both for the development of proofs or programs, these tools
attempt to cope with the fact that most of a mathematician or
programmer's time is actually spent \emph{editing}, not
\emph{writing}.

As intelligent and widely adopted as this tool chain has gotten, we
believe that they are not adapted for the new, demanding requirements
of proof developments. Indeed, whereas compilation of a program is
usually fast enough for the programmer to rely on the usual
interaction loop ((edit; compile)*; commit)*, the operation of proof
checking is usually too expensive computationally to mimic this
workflow. But even besides the time factor, this ``traditional'' way
of formalizing mathematics hinders the process of mathematical
discovery process: once a concept contained in a file is compiled, it
is considered frozen and any changes to it require the recompilation
of the whole file; the linearity of the development also gives no room
for alternate, inequivalent definitions. This fact has nonetheless
been shown to be of crucial to the mathematical discovery
process~\cite{lakatos1964proofs}, and we believe that they should be
taken into account in the realization of mathematical assistants.

@@

We propose to discuss a small part of these questions, namely the
enhancement and adaptation of version control paradigms to the
management of mathematical repositories, to witness with more
precision the \emph{impact of changes}. In proof assistants based on the
propositions-as-types paradigm, Type Theory offers a powerful tool to
witness the effect of changes finely: types.

@@

\section{A kernel language to describe semantic repositories}
\label{kernel}

@@

\remplan{Fri Jun 18, 2010 10:09 AM. Pour commencer, on peut essayer 
  de répondre à la question: ``À quoi sert ce langage?''. }

\remtext{Fri Jun 18, 2010 10:10 AM. Il sert à décrire des objets d'une
  théorie et, surtout, des transformations de ces objets (valides) en
  de nouveaux objets (valides). Il sert donc à décrire des applications
  de méta-théorèmes.}

\remplan{Fri Jun 18, 2010 10:13 AM. Ensuite, est-ce qu'il n'existe pas déjà
  des langages adaptés à ça? }

\remtext{Fri Jun 18, 2010 10:15 AM. Pas tout à fait. Certes, LF
  fournit un cadre qui autorise la représentation des metathéories et
  des objets, construits par application de règles
  admissibles/métathéorèmes. Cependant, la syntaxe de ces objets 
  est statique/extensionnel~: on décrit l'objet comme le résultat
  de sa construction en oubliant son histoire. Pour représenter 
  l'historique d'un développement, notre langage
  doit au contraire maintenir l'ordre dans lequel les objets ont
  été construits ainsi que la façon dont ils l'ont été.}

\remtext{Fri Jun 18, 2010 10:26 AM. Ce n'est pas peut être pas tout à fait vrai que personne n'a
  fait de proposition de langages qui décrivent un développement ...
  On peut peut-être citer Pfenning, Klint and co et faire une référence
  en avant vers le related work.}

\remtext{Fri Jun 18, 2010 10:20 AM. Remarque annexe: je pense qu'à ce
  stade de l'explication, il n'est pas nécessaire de parler
  d'incrémentalité et de partage maximal. Je crois que c'est une
  conséquence pratique heureuse de notre formalisation mais qui ne
  fait pas réellement partie des nécessités/besoins initiaux.
  Ce que je veux dire, c'est que l'incrémentalité est une propriété
  orthogonale à la capacité de dénoter une historique.}

Representing syntax and logics is nicely done in a \emph{logical
  framework} like LF (\cite{harper1993framework}): both the syntactic
elements and the typing derivations can sit in the same tree
structure, and both can be rechecked at the same time, thanks to
dependent types.

For the purpose of incremental type checking though, our needs are a
little bit different: first, we need to record, that is to name all
intermediate values of our developments, so as to be able to address
and reuse them multiple times. Secondly, we need to make sure that
those intermediate values (sub-terms) are not recorded twice, so as to
not type-check them twice. Thus we are looking to represent syntactic
and typing objects as a \emph{DAG} rather than a tree, and ensure
\emph{maximal sharing} among its sub-DAGs.

@@

\remtext{Fri Jun 18, 2010 10:26 AM. 
  Du coup, il faudrait reformuler un peu les paragraphes précédents.}

>>>>

Practical efforts in the formalization of mathematical results have
naturally led to the questions of how to manage large repositories of
proofs, i.e. what is the \emph{daily workflow} of users of a proof
assistant. How does one elaborate a formal proof? What kind of \emph{a
  posteriori} modification is he prone to doing? What do these
modifications imply on the validity of the whole edifice or, the other
way around, how do one rely on existing work to build up new results?
Many of these questions remain largely unanswered, but the tendency
seems to be to adapt existing methods coming from software
development, as illustrated for example by the introduction of
modules, file-based scripts and separate compilation in proof
assistants like \textsf{Coq}~{\cite{CoqDocWeb}} or
\textsf{Matita}~{\cite{Asperti06userinteraction}}, the use of
dependency management tools (\textsf{Make}) or version control system
(\textsf{GIT}, \textsf{Subversion}) to build and manage versions of a
project. Both for the development of proofs or programs, these tools
attempt to cope with the fact that most of a mathematician or
programmer's time is actually spent \emph{editing}, not
\emph{writing}.

We believe that these tools are not adapted for the new, demanding
requirements of proof developments. Indeed, whereas compilation of a
program is usually fast enough for the programmer to rely on the usual
interaction loop ((edit; compile)*; commit)*, the operation of proof
checking is usually too expensive computationally to mimic this
workflow. But even besides the time factor, this ``traditional'' way
of formalizing mathematics hinders the process of mathematical
discovery process: once a concept contained in a file is compiled, it
is considered frozen and any changes to it require the recompilation
of the whole file; the linearity of the development also gives no room
for alternate, inequivalent definitions. This fact has nonetheless
been shown to be of crucial to the mathematical discovery
process~\cite{lakatos1964proofs}, and we believe that they should be
taken into account in the realization of mathematical assistants.

@@

We propose to discuss a small part of these questions, namely the
enhancement and adaptation of version control paradigms to the
management of mathematical repositories, to witness with more
precision the \emph{impact of changes}. In proof assistants based on
the propositions-as-types paradigm, Type Theory offers a powerful tool
to witness the effect of changes finely: types, and we base our work
on these.

@@

\section{A core language to describe typed repositories}
\label{kernel}

The kernel of our system is a type-checker algorithm for a typed
meta-language. In this language, we will declare both the syntax of
the object (proof-)language and its typing rules, and define pieces of
syntax (our proofs) and their derivations. Describing transformations
among syntax objects is done by sharing common subterms or
subderivations.

@@

Representing syntax and logics is nicely done in a \emph{logical
  framework} like LF (\cite{harper1993framework}): both the syntactic
elements and the typing derivations can sit in the same tree
structure, and both can be rechecked at the same time, thanks to
dependent types. For the purpose of incremental type checking though,
our needs are a little bit different: first, we need to record, that
is to name all intermediate values of our developments, so as to be
able to address and reuse them multiple times. Secondly, we need to
make sure that those intermediate values (sub-terms) are not recorded
twice, so as to not type-check them twice: we are looking to represent
syntactic and typing objects as a DAG rather than a tree.

@@
